\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{colorlinks=true,linkcolor=blue,urlcolor=blue}

\newcommand{\vu}{\bm{u}}
\newcommand{\vf}{\bm{f}}
\newcommand{\vx}{\bm{x}}
\newcommand{\vk}{\bm{k}}
\newcommand{\vr}{\bm{r}}
\newcommand{\grad}{\nabla}
\newcommand{\dt}{\Delta t}
\newcommand{\dx}{\Delta x}

\title{\textbf{CellFlow: Technical Manual}\\[0.3em]
\large A hydrodynamically coupled agent-based cell biophysics simulator}
\author{CellFlow project}
\date{\today}

\begin{document}
\maketitle
\tableofcontents
\bigskip

\section{Overview}

CellFlow is a two-dimensional, agent-based simulator that couples discrete
cells to continuous chemical fields and to a low-Reynolds-number fluid. Each
cell is an autonomous agent that grows, divides, dies, consumes and secretes
chemicals, and interacts mechanically with its neighbours. The distinguishing
feature of CellFlow is a \emph{two-way} coupling to the fluid: cells exert
forces on the fluid, the fluid transports the chemical fields and moves the
cells, cells \emph{sense} the local fluid shear (mechanotransduction), and cells
\emph{remodel} the fluid's permeability by depositing extracellular matrix.

The state of the system at time $t$ consists of
\begin{itemize}
  \item a population of cells $\{c_i\}$, each with position $\vx_i$, radius
        $R_i$, accumulated nutrient $n_i$, polarity $\phi_i$, deviatoric shape
        strain $\bm{\varepsilon}_i$, cell type $\tau_i$, and phase;
  \item scalar fields on a uniform grid: nutrient $N(\vx)$, attractant
        $A(\vx)$, and extracellular matrix density $E(\vx)$;
  \item the fluid velocity field $\vu(\vx)$ and drag field $\alpha(\vx)$.
\end{itemize}

Throughout, the domain is the square $[0,L]^2$ discretized into a
$G\times G$ grid with spacing $\dx = L/G$. Time advances in steps of size $\dt$.

\section{The simulation step}\label{sec:step}

One time step executes the following operations in order
(see \texttt{CellSimulation.\_simulation\_step}):
\begin{enumerate}
  \item Assemble cell arrays (positions, radii).
  \item Compute forces: chemotactic propulsion, adhesion, repulsion, and
        post-division forces (Section~\ref{sec:forces}).
  \item Solve the fluid velocity from the force distribution
        (Section~\ref{sec:fluid}).
  \item[3b.] Mechanotransduction: update cell polarity from the fluid strain
        rate (Section~\ref{sec:mechano}).
  \item[3c.] Cell-shape mechanics: update deviatoric strain from contact stress
        (Section~\ref{sec:shape}).
  \item Check the advective CFL condition.
  \item Advect the scalar fields with the fluid velocity.
  \item Diffuse the scalar fields.
  \item Cell biology: nutrient uptake, secretion, metabolism, growth, phase,
        death (Section~\ref{sec:biology}).
  \item Move cells with the mobility relation (Section~\ref{sec:motion}).
  \item Handle division and death; resolve overlaps; enforce boundaries.
\end{enumerate}
The fluid is quasi-static (Stokes/Brinkman): it is recomputed each step from the
instantaneous force distribution, with no time history.

\section{Cell biology}\label{sec:biology}

\subsection{Nutrient and growth}
Each cell consumes nutrient over its circular footprint at a per-cell rate
$k_i \sim \mathcal{N}(0.2, 0.05)$, accumulating internal nutrient $n_i$. Per step
the cell takes up $\Delta n_i^{+}$ (the integrated uptake over its area, capped
by the locally available nutrient), pays a basal metabolic cost
$b\,\dt$ with $b=0.02$, and optionally secretes attractant at rate $s=1.0$:
\begin{equation}
  n_i \leftarrow n_i + \Delta n_i^{+} - b\,\dt .
\end{equation}
A cell dies (is removed) when $n_i < 0$.

\paragraph{Uptake kinetics.} The local uptake rate as a function of the ambient
concentration $c$ is selectable. By default it is \emph{first-order} (linear),
$q(c)=k_i\,c$. Setting a half-saturation constant $K_m$ (per cell, or globally
via \texttt{nutrient\_uptake\_saturation}) switches to \emph{Michaelis--Menten /
Monod} saturating kinetics,
\begin{equation}
  q(c) = k_i\,c\,\frac{K_m}{K_m + c},
\end{equation}
so that $k_i$ is the low-concentration rate constant ($q\to k_i c$ as
$c\ll K_m$) and the uptake saturates at $k_i K_m$ for $c\gg K_m$ (a cell cannot
consume arbitrarily fast). The linear law is recovered exactly as
$K_m\to\infty$. Saturating uptake reshapes the nutrient profile and is the
realistic regime for diffusion-limited aggregates and colony fronts.

\subsection{Size models}
CellFlow provides two growth laws, selected by \texttt{growth\_model}.

\paragraph{Linear (legacy).} The radius is linear in accumulated nutrient,
\begin{equation}
  R_i = \mathrm{clip}\!\left(R_{\min} + n_i\,\frac{R_{\max}-R_{\min}}{100},\,
  R_{\min},\, R_{\max}\right),
\end{equation}
with $R_{\min}=2$, $R_{\max}=4$. Because area $\propto R^2$, area grows
quadratically in $n_i$, and division (below) is \emph{not} area-conserving.

\paragraph{Area-conserving (recommended for physical fidelity).} The cell
\emph{area} grows linearly with accumulated nutrient (mass), and the radius is
recovered from the area:
\begin{equation}
  A_i = \pi R_{\max}^2 \,\frac{n_i}{100}, \qquad
  R_i = \min\!\left(\sqrt{A_i/\pi},\, R_{\max}\right).
\end{equation}
At the division threshold ($n_i=100$, $R_i=R_{\max}$), halving $n_i$ halves the
area, so each of the two daughters has area $A_{\max}/2$ and radius
$R_{\max}/\sqrt{2}$; the two daughters therefore conserve the mother's area
exactly.

\subsection{Division and death}
A cell enters the \texttt{DIVISION} phase when $R_i \ge R_{\max}$. On division it
halves its accumulated nutrient, returns to \texttt{GROWTH}, and creates a
daughter placed at contact distance $R_{\text{parent}} + R_{\text{daughter}}$
along a random unit direction (so the daughter is born just touching, not
overlapping). The daughter inherits cell type, polarity, and growth mode; its
deviatoric shape strain starts at zero (a newborn is round). For a brief window
after division the mother--daughter pair feel a mutual separating force.

\section{Chemical fields}\label{sec:fields}

Each scalar field $c \in \{N, A\}$ obeys advection--diffusion with cellular
source/sink terms,
\begin{equation}
  \frac{\partial c}{\partial t} + \vu\cdot\grad c
  = D_c \grad^2 c + S_c(\vx,t),
\end{equation}
solved by operator splitting each step: advection, then diffusion, then the
cellular uptake/secretion. Boundary conditions for the nutrient field are
selectable: \emph{Neumann} (no-flux, isolated system) or \emph{Dirichlet}
(edges held at a constant reservoir value, modelling an external medium).

\paragraph{Diffusion.} Two solvers are available, selected by
\texttt{diffusion\_solver}. The default \emph{explicit} five-point (FTCS) scheme,
\begin{equation}
  c^{m+1}_{ij} = c^{m}_{ij}
  + \frac{D_c \dt}{\dx^2}\left(c_{i-1,j}+c_{i+1,j}+c_{i,j-1}+c_{i,j+1}-4c_{ij}\right),
\end{equation}
clamped to be non-negative, is stable only when the diffusion number
$D_c\dt/\dx^2 \le 1/4$ (the code warns otherwise).

The \emph{implicit} solver (\texttt{diffusion\_solver='implicit'}) is an
alternating-direction implicit (ADI, Peaceman--Rachford) scheme: each step is two
half-steps, the first implicit in $x$ and explicit in $y$, the second the reverse,
\begin{align}
  \big(\mathcal{I} - \tfrac{r}{2}\,\delta_x^2\big)c^{*}
   &= \big(\mathcal{I} + \tfrac{r}{2}\,\delta_y^2\big)c^{m}, \\
  \big(\mathcal{I} - \tfrac{r}{2}\,\delta_y^2\big)c^{m+1}
   &= \big(\mathcal{I} + \tfrac{r}{2}\,\delta_x^2\big)c^{*},
\end{align}
with $r = D_c\dt/\dx^2$ and $\delta^2$ the second-difference operator. Each
half-step is a set of independent tridiagonal systems solved by the Thomas
algorithm ($O(M)$), parallelized over lines. ADI is \emph{unconditionally stable}
(any $\dt$), second order in space and time, and preserves the discrete steady
state exactly (a uniform field at the Dirichlet value is a fixed point), so it
reaches the quasi-steady profile in far fewer steps than the explicit scheme and
without a CFL limit on $D$. Because CellFlow advects fields by the flow, the
overall scheme is IMEX (implicit diffusion combined with the semi-Lagrangian
advection below), which keeps the cell--fluid coupling intact. ADI is
$L^2$-stable rather than strictly maximum-principle-preserving, so very large
steps can produce small ($\sim$few-percent) over/undershoots, and with a fixed
step it converges to the exact steady state gradually (no parameter cycling).

Under no-flux (Neumann) boundaries the explicit scheme is conservative in the
interior (verified) but the node-centred ghost treatment is not strictly
flux-conservative at the walls.

\paragraph{Advection.} A semi-Lagrangian scheme back-traces each grid point
along $-\vu\,\dt$ and interpolates bilinearly. It is unconditionally stable but
diffusive and first-order; it conserves mass in the interior under smooth flow
(verified to $\sim 10^{-5}$) but loses mass where material is advected into the
clamped domain boundary.

\section{Cell forces}\label{sec:forces}

Forces on the cells are separated into a \emph{propulsion} component (active,
chemotactic) and a \emph{monopolar} component (passive, mechanical), so that
the dipole hydrodynamic model can treat propulsion as force-free.

\paragraph{Chemotactic propulsion.} Each cell samples the area-averaged nutrient
gradient and combines it with a random walk to set a direction, then propels at
fixed magnitude $F_{\text{prop}}$:
\begin{equation}
  \hat{\vf}_i \propto \chi\,\overline{\grad N}_i + w\,\bm{\eta}_i, \qquad
  \vf^{\text{prop}}_i = F_{\text{prop}}\,\frac{\hat{\vf}_i}{\lvert\hat{\vf}_i\rvert},
\end{equation}
where $\chi$ is the chemotactic coefficient, $w$ the random-walk speed, and
$\bm{\eta}_i$ standard normal noise. The noise is drawn once per step in the main
thread (seedable) and passed into the parallel kernel, so results are
reproducible and independent of thread count.

\paragraph{Repulsion (volume exclusion).} For overlapping cells
($d_{ij} < R_i + R_j$), with overlap $\delta_{ij}=R_i+R_j-d_{ij}$,
\begin{equation}
  \vf^{\text{rep}}_i = -\sum_{j} k_{\text{rep}}\,
  \exp\!\left(\frac{3\delta_{ij}}{R_i+R_j}\right)\hat{\vr}_{ij},
\end{equation}
directed away from the neighbour.

\paragraph{Adhesion.} For cells within an outer band
$R_i+R_j < d_{ij} < \lambda(R_i+R_j)$ (cutoff factor $\lambda$), an attractive
force linear in the separation beyond contact,
\begin{equation}
  \vf^{\text{adh}}_i = \sum_j k_{\text{adh}}\,(d_{ij}-(R_i+R_j))\,\hat{\vr}_{ij}.
\end{equation}

\paragraph{Differential adhesion.} With cell types and a symmetric adhesion
matrix $\bm{D}$, the scalar strength $k_{\text{adh}}$ is replaced by
$D_{\tau_i\tau_j}$. Following the Differential Adhesion Hypothesis, the
interfacial tension between types is
$\sigma_{ab} = D_{ab} - \tfrac{1}{2}(D_{aa}+D_{bb})$, which drives cell sorting
and engulfment. A uniform matrix reproduces the scalar law exactly.

\paragraph{Neighbour lists.} All pairwise force loops use a uniform linked-cell
(bin) structure when enabled (default), reducing the cost from $O(N^2)$ to
$O(N)$; the binned forces are identical to the brute-force result to
$\sim 10^{-10}$.

\paragraph{Overlap resolution.} After the force-driven move, residual overlaps
are removed by a geometric projection: each overlapping pair is pushed apart by
half the overlap, iterated a configurable number of times per step. With
neighbour lists enabled (default) this uses a parallel cell-list \emph{Jacobi}
sweep---each cell accumulates the half-overlap push from its overlapping
neighbours (read-only on positions) and all displacements are applied at
once---which is $O(N)$ rather than the $O(N^2)$ sequential (Gauss--Seidel)
sweep, and preserves the centre of mass exactly (equal-and-opposite pushes);
an isolated pair is resolved in a single sweep. The degenerate
exactly-coincident case is separated along a deterministic golden-angle
direction, preserving reproducibility.

\section{Hydrodynamics}\label{sec:fluid}

The fluid is governed by quasi-static, incompressible low-Reynolds flow. Three
solvers are available, selected by \texttt{fluid\_model} and related options.

\subsection{Free-space regularized Stokeslet (legacy)}
The velocity is the sum of 2D regularized Stokeslets (Cortez) over the cell
forces. For a force $\vf$ at $\vx_0$, with $\vr=\vx-\vx_0$,
$r_\epsilon^2 = \lvert\vr\rvert^2+\epsilon^2$ and regularization
$\epsilon = 2\dx$,
\begin{equation}
  \vu(\vx) = \frac{1}{4\pi\mu}\Big[
  -\vf\big(\ln r_\epsilon - \tfrac{\epsilon^2}{r_\epsilon^2}\big)
  + (\vf\cdot\vr)\,\frac{\vr}{r_\epsilon^2}\Big].
\end{equation}
A force-dipole variant represents active propulsion as two equal-and-opposite
Stokeslets (no net force on the fluid). \textbf{Caveat:} in 2D the Stokeslet
velocity grows like $\ln r$ for a net force (the Stokes paradox), so this model
has no far-field decay and no confinement; it is retained for backward
compatibility and force-free (dipole) studies.

\subsection{FFT Brinkman solver (recommended)}
The preferred model solves the incompressible Brinkman equations on a periodic
grid,
\begin{equation}
  -\mu\grad^2\vu + \alpha\,\vu + \grad p = \vf, \qquad \grad\cdot\vu = 0,
  \label{eq:brinkman}
\end{equation}
where $\mu$ is the viscosity and $\alpha$ the substrate drag. The drag
introduces a screening length $\delta = \sqrt{\mu/\alpha}$ beyond which
hydrodynamic interactions decay; this regularizes the 2D paradox. In Fourier
space, for each wavevector $\vk\neq 0$,
\begin{equation}
  \hat{\vu}(\vk) = \frac{\mathsf{P}(\vk)\,\hat{\vf}(\vk)}{\mu\lvert\vk\rvert^2+\alpha},
  \qquad
  \mathsf{P}(\vk) = \mathsf{I} - \frac{\vk\vk^{\mathsf{T}}}{\lvert\vk\rvert^2},
  \label{eq:brinkman-fft}
\end{equation}
where $\mathsf{P}$ is the Leray (divergence-free) projection. The mean mode is
$\hat{\vu}(0)=\hat{\vf}(0)/\alpha$ (finite for $\alpha>0$). The cost is
$O(M\log M)$ for $M=G^2$ grid points. The Nyquist wavenumber is zeroed on even
grids so the projection is consistent (otherwise the velocity is not
divergence-free). The solver is verified against an analytic manufactured
solution to machine precision and is divergence-free to $\sim 10^{-9}$.

\subsection{Free-slip box (non-periodic walls)}\label{sec:freeslip}
Setting \texttt{fluid\_boundary='freeslip\_box'} replaces the periodic boundaries
with free-slip (no-penetration, stress-free) walls on all four sides of the
square domain,
\begin{equation}
  \vu\cdot\bm{n} = 0, \qquad
  \big(\mathsf{\sigma}\bm{n}\big)\cdot\bm{t} = 0
  \quad\text{on each wall.}
\end{equation}
A free-slip wall is a plane of mirror symmetry, so the box solution is the
restriction of a periodic solution on the doubled domain $[0,2L]^2$ whose force
is extended with the parity $f_x$ odd across the $x$-walls and even across the
$y$-walls, and $f_y$ even across the $x$-walls and odd across the $y$-walls (the
extensions underlying the discrete sine/cosine transforms). The Brinkman
multiplier in \eqref{eq:brinkman-fft} and the Leray projection $\mathsf{P}$ both
map this symmetry subspace to itself, so the solved velocity inherits the
symmetry---$u_x$ odd in $x$ and even in $y$, $u_y$ even in $x$ and odd in
$y$---which is exactly free slip, and remains divergence-free. Reusing the
verified periodic solver on the mirror-extended grid keeps the cost
$O(M\log M)$ (a constant factor of $4$). This path is verified against a
manufactured free-slip mode to machine precision. It is currently restricted to
constant drag $\alpha$ (not combined with the poroelastic ECM solver), and no-slip
walls would instead require a Chebyshev/influence-matrix formulation.

\subsection{Variable-coefficient (poroelastic) Brinkman}
When cells remodel the matrix (Section~\ref{sec:ecm}), the drag becomes a field
$\alpha(\vx)$, and \eqref{eq:brinkman} is no longer diagonal in Fourier. Writing
$\alpha(\vx) = \alpha_0 + d(\vx)$ with $\alpha_0 = \max_{\vx} \alpha(\vx)$ and
$d \le 0$, the solution satisfies the fixed point
\begin{equation}
  \vu = \mathsf{A}_0^{-1}\mathsf{P}\big(\vf - d\,\vu\big),
\end{equation}
where $\mathsf{A}_0^{-1}\mathsf{P}$ is precisely the constant-$\alpha_0$ FFT
solve \eqref{eq:brinkman-fft}. This Picard iteration contracts because
$\lvert d\rvert/(\mu\lvert\vk\rvert^2+\alpha_0) < 1$, with rate
$\sim 1 - \alpha_{\min}/\alpha_{\max}$. It is efficient for moderate drag
contrast; very high contrast would warrant a preconditioned Krylov solver
(future work). It is verified by constant-$\alpha$ consistency, a
spatially-varying manufactured solution (to $10^{-6}$), and flow screening
through a high-drag patch.

\subsection{Immersed Boundary coupling}\label{sec:ibm}
Cells couple to the grid fluid by the Immersed Boundary Method. A cell force
$\vf_i$ is \emph{spread} to a grid force density with a per-cell Gaussian blob of
\emph{physical} width $\sigma_i = \zeta R_i$ (where $\zeta$ is
\texttt{ibm\_reg\_factor}):
\begin{equation}
  \vf(\vx) = \sum_i \vf_i\,\varphi_{\sigma_i}(\vx-\vx_i),
\end{equation}
normalized per cell so that total force is conserved exactly. The solved
velocity is \emph{interpolated} back to each cell with the same kernel,
\begin{equation}
  \vu_i = \sum_{\vx} \vu(\vx)\,\hat{\varphi}_{\sigma_i}(\vx-\vx_i),
\end{equation}
with weights summing to one (partition of unity), so a uniform field is
recovered exactly. Tying the regularization to the cell radius (rather than the
grid spacing, as a grid-width kernel would) makes the single-cell self-mobility
\emph{grid convergent}: it approaches a finite limit as $\dx\to 0$, whereas a
grid-tied kernel drifts (logarithmically). The same field $\vu$ both advects the
scalar fields and moves the cells, removing any kinematic inconsistency.

\subsection{Cell motion}\label{sec:motion}
In the Brinkman/IBM path, cell velocities are the interpolated fluid velocity at
the cell centres. In the legacy Stokeslet path, the velocity is a mobility
relation combining a self term (Stokes drag) and hydrodynamic interactions:
\begin{equation}
  \vu_k = \frac{\vf_k}{6\pi\mu R_k}
  + \sum_{j\neq k} \mathsf{G}_{2D}(\vx_k,\vx_j)\,\vf_j .
\end{equation}
Cells advance by forward Euler, $\vx_k \leftarrow \vx_k + \vu_k\,\dt$, which is
first order in $\dt$ (verified).

\section{Mechanotransduction}\label{sec:mechano}

Cells sense the local fluid shear and reorient. From the solved velocity, the
velocity-gradient tensor $\grad\vu$ is computed by periodic central differences
and area-averaged over each cell. The strain-rate tensor is
$\mathsf{E} = \tfrac{1}{2}(\grad\vu + \grad\vu^{\mathsf{T}})$, from which we
form the scalar shear rate and the principal (extensional) axis,
\begin{align}
  \dot\gamma &= \sqrt{2\,\mathsf{E}\!:\!\mathsf{E}}
  = \sqrt{2\big(E_{xx}^2+E_{yy}^2+2E_{xy}^2\big)}, \\
  \psi &= \tfrac{1}{2}\,\mathrm{atan2}\big(2E_{xy},\, E_{xx}-E_{yy}\big).
\end{align}
The cell polarity $\phi$ (a nematic director, defined modulo $\pi$) aligns toward
the strain axis at a rate proportional to the shear,
\begin{equation}
  \frac{d\phi}{dt} = -k_\phi\,\dot\gamma\,\sin\!\big(2(\phi-\psi)\big),
\end{equation}
with rate constant $k_\phi$ (\texttt{shear\_alignment\_rate}). The stable fixed
point is $\phi=\psi$. Optionally the polarity drives an active propulsion of
magnitude \texttt{polarity\_propulsion\_force} (rheotaxis). Verification: under
analytic simple shear $u=(\dot\gamma\,y,0)$ the sensed shear rate equals
$\dot\gamma$, the axis is $45^\circ$, and the polarity converges to $45^\circ$;
zero flow leaves the polarity unchanged.

\section{Cell-shape mechanics}\label{sec:shape}

Cells deform into area-conserving ellipses under contact pressure, while the
mechanical contacts remain circular (so the cost stays low). For each cell the
repulsive contact (Cauchy) stress is
\begin{equation}
  \bm{\sigma}_i = \frac{1}{A_i}\sum_j \vf_{ij}\otimes\vr_{ij},
\end{equation}
with deviatoric part
$\bm{s}_i = \big(\tfrac{1}{2}(\sigma_{xx}-\sigma_{yy}),\, \sigma_{xy}\big)$. The
deviatoric strain $\bm{\varepsilon}_i=(\varepsilon_{xx},\varepsilon_{xy})$
relaxes by a linear viscoelastic law toward $\chi\,\bm{s}_i$,
\begin{equation}
  \bm{\varepsilon}_i \leftarrow \bm{\varepsilon}_i
  + \big(\chi\,\bm{s}_i - \bm{\varepsilon}_i\big)\frac{\dt}{\tau},
\end{equation}
capped at a maximum aspect ratio, with compliance $\chi$
(\texttt{shape\_compliance}) and relaxation time $\tau$
(\texttt{shape\_relaxation\_time}). The ellipse semi-axes follow from the strain
magnitude $m=\lvert\bm{\varepsilon}_i\rvert$,
\begin{equation}
  a = R_i\,e^{m}, \quad b = R_i\,e^{-m}, \quad
  \theta = \tfrac{1}{2}\mathrm{atan2}(\varepsilon_{xy},\varepsilon_{xx}),
\end{equation}
so that $ab=R_i^2$ (area conserved). Verification: uniaxial compression
elongates perpendicular to the squeeze and reaches the elastic steady state
$\bm{\varepsilon}=\chi\bm{s}$; \emph{isotropic} (hydrostatic) pressure produces
\emph{no} elongation; the shape relaxes back to round when unloaded.

\section{Dynamic extracellular matrix}\label{sec:ecm}

Cells deposit ECM into a grid field $E(\vx)$ over their area; the field decays
and is mapped to the local Brinkman drag,
\begin{equation}
  E \leftarrow (1 - \kappa_d\,\dt)\,E + \text{(cell secretion)}, \qquad
  \alpha(\vx) = \alpha_{\text{base}} + \kappa_\alpha\,E(\vx),
\end{equation}
with secretion rate, decay rate $\kappa_d$, and drag coefficient $\kappa_\alpha$
set by \texttt{ecm\_secretion\_rate}, \texttt{ecm\_decay\_rate}, and
\texttt{ecm\_drag\_coeff}. The resulting variable drag field is passed to the
poroelastic solver, so cell-deposited matrix lowers the local permeability and
reroutes/screens the flow---the defining cell$\to$fluid feedback (relevant to
tumour stroma, interstitial flow, and durotaxis).

\section{Numerical methods and stability}

\begin{itemize}
  \item \textbf{Time integration:} forward Euler for cell positions and the
        ODE-like polarity/shape relaxations (first order in $\dt$).
  \item \textbf{Diffusion:} selectable --- explicit FTCS (stable for
        $D\dt/\dx^2 \le 1/4$) or implicit ADI (Peaceman--Rachford, Thomas solves,
        unconditionally stable, second order, IMEX with the flow advection).
  \item \textbf{Uptake:} first-order (linear) or Michaelis--Menten/Monod
        saturating kinetics.
  \item \textbf{Advection:} semi-Lagrangian; the code warns when the CFL number
        $\max\lvert\vu\rvert\,\dt/\dx > 1$.
  \item \textbf{Fluid:} spectral (FFT) for Brinkman, $O(M\log M)$; Picard
        iteration for variable drag.
  \item \textbf{Pairwise forces:} linked-cell neighbour lists, $O(N)$.
\end{itemize}

\section{Verification and validation}\label{sec:vv}

CellFlow ships an automated test suite (115 tests) including:
\begin{itemize}
  \item \textbf{Solver verification:} a Method of Manufactured Solutions test
        recovers a multi-mode divergence-free field to machine precision for
        both constant and spatially-varying drag.
  \item \textbf{Convergence:} grid refinement studies confirm that the IBM
        self-mobility and hydrodynamic interactions converge; the timestep error
        scales as $\dt^1$ (forward Euler); the advection scheme conserves mass
        in the interior.
  \item \textbf{Transport solvers:} the implicit ADI diffusion is verified
        unconditionally stable at $\dt$ far beyond the explicit CFL limit, exact
        on the uniform steady state, mass-conserving under Neumann, and in
        agreement with the explicit scheme over a transient at small $\dt$.
  \item \textbf{Physics checks:} screening monotonicity; mechanotransduction
        alignment to the correct strain axis; cell-shape response to uniaxial vs
        isotropic stress; area-conserving division; Michaelis--Menten uptake
        (reduces uptake at high concentration, recovers the linear law as
        $K_m\to\infty$, saturates, stays non-negative).
  \item \textbf{Reproducibility:} with a seed, runs are bit-for-bit identical
        and independent of the thread count.
\end{itemize}
\emph{Verification} (solving the equations correctly) is extensive;
\emph{validation} against experimental data is the user's responsibility for any
quantitative study, and physical units must be assigned by the user.

\section{Limitations}

\begin{itemize}
  \item Two spatial dimensions only. Boundaries are either periodic with
        screening as a confinement surrogate, or free-slip walls on all four
        sides (Section~\ref{sec:freeslip}); explicit no-slip walls are not yet
        available.
  \item Overlap resolution is a geometric projection independent of viscosity;
        dense-tissue mechanics are therefore not viscosity-resolved.
  \item The differential-adhesion law is qualitative (not calibrated to a
        measured surface tension).
  \item The poroelastic Picard solver is efficient only for moderate drag
        contrast.
  \item Results are in arbitrary units unless the user assigns a unit system.
\end{itemize}

\begin{thebibliography}{9}
\bibitem{cortez} R.~Cortez, \emph{The method of regularized Stokeslets},
  SIAM J. Sci. Comput. 23 (2001) 1204--1225.
\bibitem{peskin} C.~S.~Peskin, \emph{The immersed boundary method},
  Acta Numerica 11 (2002) 479--517.
\bibitem{brinkman} H.~C.~Brinkman, \emph{A calculation of the viscous force
  exerted by a flowing fluid on a dense swarm of particles},
  Appl. Sci. Res. A1 (1949) 27--34.
\bibitem{steinberg} M.~S.~Steinberg, \emph{Differential adhesion in
  morphogenesis}, Curr. Opin. Genet. Dev. 17 (2007) 281--286.
\end{thebibliography}

\end{document}
